---
title: "ICI_fecal_metabolome"
author: "Ruben Mars"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  
  read_MannWhitney_fun: "helper functions/MannWhitney_function.R"
  read_boxplot_fun: "helper functions/make_boxplot_figure.R"
  replace_outliers_with_median: "helper functions/replace_outliers_with_median.R"
  
  MEGA_w_HMDB: "../data/TMIC MEGA data template with HMDB IDs.xlsx"
  HMDB_pathways: "helper functions/plasma mx enricher/annotation_list_hmdb_all_Sept2018.csv"  

output: 
  html_document:
    toc: true
    theme: cerulean
    
---

<style type="text/css">
.main-container {
  max-width: 1200px;
  margin-left: auto;
  margin-right: auto;
}
</style>

***  

## Content


Analysis on pretreatment gut microbiome sequencing and metabolomics data on cohort of cancer patients treated with ICI



***  


```{r, echo=FALSE, results='hide',message=FALSE}

require(tidyverse)
require(readxl)
require(ggplot2)
require(RColorBrewer)
require(vegan)
require(ape)
require(tidyverse)
require(ggpubr)
require(cowplot)
require(randomForest)
require(ROCR)
require(gtsummary)
require(flextable)
require(gridExtra)
library(ggpubr)
require(reshape2)
require(dplyr)
require(openxlsx)
require(knitr)
#library(devtools)
#install_github("jokergoo/ComplexHeatmap")
require(ComplexHeatmap)
require(circlize)
require(mlbench)
require(caret)
#BiocManager::install("ropls")
require(ropls)
require(GUniFrac)
require(GenericML)
library(dplyr)
library(tibble)
library(ggrepel)
require(vegan)
require(outliers)


source(params$read_MannWhitney_fun)
source(params$read_boxplot_fun)
source(params$replace_outliers_with_median)

source("helper functions/zicoseq_volcano_plot.R")
source("helper functions/zicoseq_heatmap_old_colors.R")
source("helper functions/plasma mx enricher/enricher_function.R")


```


Run rmarkdown file to load the metadata

```{r, echo=FALSE, results='hide',message=FALSE, warning=FALSE}

rmarkdown::render("1_ICI_cohort_description.Rmd")

rownames(clin_meta) <- clin_meta$SampleID

clin_meta_mx <- clin_meta[!is.na(clin_meta$Stool_metabolomics_SampleID),]
rownames(clin_meta_mx) <- clin_meta_mx$SubjectID


```


Load metabolomics data

```{r, echo=FALSE, results='hide',message=FALSE}

filename <- "../data/Supplementary Table 1.xlsx"
#getSheetNames(filename)
mx_data <- as.data.frame(read.xlsx(filename, sheet = "stool_mx", rowNames = T))


```

Add pathway level; remove lipids

```{r, echo=FALSE, results='hide',message=FALSE}

#Get compound classes for pathway enrichment / filtering
HMDB_pathways <- as.data.frame(read.csv(params$HMDB_pathways))
head(HMDB_pathways)
rownames(HMDB_pathways) <- HMDB_pathways$hmdb_id

MEGA_w_HMDB <- as.data.frame(read.xlsx(params$MEGA_w_HMDB, sheet=1, rowNames = F))
MEGA_w_HMDB$metabolite.name <- gsub(" ", "\\.", MEGA_w_HMDB$metabolite.name)

table(MEGA_w_HMDB$HMDB.IDs %in% HMDB_pathways$hmdb_id) #matching by ID is better

#match 
table(rownames(mx_data) %in% MEGA_w_HMDB$metabolite.name)
#FALSE  TRUE 
#    7   556  

metabolite_names_for_matching <- rownames(mx_data)
metabolite_hmdb_for_matching <- as.character(sapply(metabolite_names_for_matching, function(x) MEGA_w_HMDB$HMDB.IDs[MEGA_w_HMDB$metabolite.name == x]))

#add the pathways to here
HMDB_pathways_matched <- HMDB_pathways[metabolite_hmdb_for_matching,]
#cbind(HMDB_pathways_matched$name, metabolite_names_for_matching)

rownames(HMDB_pathways_matched) <- metabolite_names_for_matching
HMDB_pathways_matched["Orotic.acid",]

dim(mx_data); dim(HMDB_pathways_matched)
#[1] 563 167
#[1] 563  12

```


Filtering metabolites based on annotation / pathway

```{r, echo=FALSE, results='hide',message=FALSE}

#head(HMDB_pathways_matched)
#write.csv(HMDB_pathways_matched, "HMDB_pathways_matched.csv")

#remove metabolites taxonomy_class NA
#optionally also remove taxonomy_subclass Fatty acid esters, Glycerophosphocholines, Phosphosphingolipids, Triradylcglycerols

#table(HMDB_pathways_matched$taxonomy_class)

HMDB_pathways_matched_sub <- HMDB_pathways_matched[!is.na(HMDB_pathways_matched$taxonomy_class),]
dim(HMDB_pathways_matched_sub)
#225  12

mx_data_sub <- mx_data[rownames(HMDB_pathways_matched_sub), ]


```


Subset and process metabolomics data

```{r, echo=FALSE, results='hide',message=FALSE}

#cbind(rownames(clin_meta_mx), rownames(mx_data))

#impute 0s
#filter - prevalence > 70%, average intensity > 1e-5; 
#imputation - half minimum; 
#normalization - Probabilistic Quotient Normalization (PQN); 
#transformation - logarithmic


#use the concentrations for beta diversity
#use this for the stats
#mtbs <- mx_data
mtbs <- mx_data_sub
dim(mtbs) 

# Filtering
hist(colMeans(mtbs != 0))
mtbs <- mtbs[rowMeans(mtbs != 0) > 0.5, ]
mtbs <- mtbs[rowMeans(mtbs) > 1e-5, ] #remove metabolites with very low abundance
dim(mtbs) #186 167
hist(colMeans(mtbs != 0))

# Imputation
mtbs <- t(apply(mtbs, 1, function (x) {x[x == 0] <- 0.5 * min(x[x != 0]); return(x)}))
mtbs_imp <- mtbs

# PQN Normalization
hist(colSums(mtbs))
ref <- apply(mtbs, 1, median)
sf <- apply(mtbs / ref, 1, median)
mtbs <- t(t(mtbs) / sf)

# log transformation
mtbs <- log2(mtbs)

hist(mtbs[sample(rownames(mtbs), 1), ])

#rownames(mtbs)
dim(mtbs)
#186 167


HMDB_pathways_matched_mtbs <- HMDB_pathways_matched_sub[rownames(mtbs), ]


```

Subset metabolomics data with clin data and put in same order

```{r, echo=FALSE, results='hide',message=FALSE}

#clin_meta_mx$water_content_perc
#cbind(rownames(clin_meta_mx), colnames(mx_data))


dim(clin_meta_mx); dim(mtbs)
#[1] 167 58
#[1] 186 167 

table(clin_meta_mx$IMDC)
#   0   1 
#  153  14

table(clin_meta_mx$hepatitis)
#  0   1 
# 153  14 

table(clin_meta_mx$pneumonitis)
#  0   1 
# 159   8 

table(clin_meta_mx$any_toxicity) 
#  0  1 
#  76 91 


```


```{r, echo=FALSE, results='hide',message=FALSE}


#Bristol_score_curated added below
variables <- c("IMDC", "hepatitis", "pneumonitis","other_irAE", "Sex", "Elixhauser_elix.sum", "Age", "abx_last_month", "BMI",
               "Bristol_score_imputed")


#run on all data, not the subsetted data
beta_div <- as.matrix(vegdist(t(mx_data), method = "manhattan"))


set.seed(1)
# Permanova - Distance based multivariate analysis of variance
uni_1a <- adonis2(beta_div ~ IMDC, data = clin_meta_mx, permutations = 999)
uni_1b <- adonis2(beta_div ~ hepatitis, data = clin_meta_mx, permutations = 999)
uni_1c <- adonis2(beta_div ~ pneumonitis, data = clin_meta_mx, permutations = 999)
uni_3 <- adonis2(beta_div ~ other_irAE, data = clin_meta_mx, permutations = 999)
uni_6 <- adonis2(beta_div ~ Sex, data = clin_meta_mx, permutations = 999)
uni_7 <- adonis2(beta_div ~ Elixhauser_elix.sum, data = clin_meta_mx, permutations = 999)
uni_8 <- adonis2(beta_div ~ Age, data = clin_meta_mx, permutations = 999)
uni_9 <- adonis2(beta_div ~ abx_last_month, data = clin_meta_mx, permutations = 999)
uni_10 <- adonis2(beta_div ~ BMI, data = clin_meta_mx, permutations = 999)
uni_11 <- adonis2(beta_div ~ Bristol_score_imputed, data = clin_meta_mx, permutations = 999)


beta_res_list <- list(uni_1a, uni_1b, uni_1c, uni_3, uni_6, uni_7, uni_8, uni_9, uni_10, uni_11)

R2_vals <- unlist(lapply(beta_res_list, function(x) x[,"R2"][1]))
p_vals <- unlist(lapply(beta_res_list, function(x) x[,"Pr(>F)"][1]))
beta_res_df <- as.data.frame(cbind(R2_vals, p_vals))

rownames(beta_res_df) <- variables

beta_res_df <- beta_res_df[order(beta_res_df$R2_vals, decreasing = F),]


#varying color for p values
#n of group size
par(mar=c(4,14,1,1))
barplot(beta_res_df$R2_vals, horiz = T, las=1, col = beta_res_df$p_vals < 0.05, names.arg = rownames(beta_res_df))
mtext("R2 % variance explained", line=2.5, side=1)


filename <- "../figures/barplot fecal mx weighted manhattan.png"
png(filename, width = 8*330, height = 5*330, res = 330)
par(mar=c(4,14,1,1))
barplot(beta_res_df$R2_vals, horiz = T, las=1, col = beta_res_df$p_vals < 0.05, names.arg = rownames(beta_res_df))
mtext("R2 % variance explained", line=2.5, side=1)
dev.off()


write.csv(beta_res_df, file = "../data/beta_res_df manhattan stool metabolome.csv")

```


Correct for water content in figure for paper

```{r, echo=FALSE, results='hide',message=FALSE}


#looks the same on all and subsetted by prevalence data
#beta_div <- as.matrix(vegdist(t(mx_data), method = "manhattan"))
beta_div <- as.matrix(vegdist(t(mx_data), method = "manhattan"))

#variables <- c("IMDC", "hepatitis", "pneumonitis",
#               "other_irAE", "any_toxicity", "treatment_success_nr_imputed", 
#               "Sex", "Elixhauser_elix.sum", "Age", "abx_last_month", "Bristol_score_curated", "BMI")

set.seed(1)
# Permanova - Distance based multivariate analysis of variance
uni_1a <- GUniFrac::adonis3(beta_div ~ Bristol_score_imputed + Age + IMDC, data = clin_meta_mx, permutations = 999)$aov.tab
uni_1b <- GUniFrac::adonis3(beta_div ~ Bristol_score_imputed + Age + hepatitis, data = clin_meta_mx, permutations = 999)$aov.tab
uni_1c <- GUniFrac::adonis3(beta_div ~ Bristol_score_imputed + Age + pneumonitis, data = clin_meta_mx, permutations = 999)$aov.tab
uni_1d <- GUniFrac::adonis3(beta_div ~ Bristol_score_imputed + Age + other_irAE, data = clin_meta_mx, permutations = 999)$aov.tab

uni_4 <- GUniFrac::adonis3(beta_div ~ Bristol_score_imputed + Age + Sex, data = clin_meta_mx, permutations = 999)$aov.tab
uni_5 <- GUniFrac::adonis3(beta_div ~ Bristol_score_imputed + Age + Elixhauser_elix.sum, data = clin_meta_mx, permutations = 999)$aov.tab
uni_7 <- GUniFrac::adonis3(beta_div ~ Bristol_score_imputed + Age + abx_last_month, data = clin_meta_mx, permutations = 999)$aov.tab
uni_9 <- GUniFrac::adonis3(beta_div ~ Bristol_score_imputed + Age + BMI, data = clin_meta_mx, permutations = 999)$aov.tab

beta_res_list <- list(uni_1a, uni_1b, uni_1c, uni_1d, uni_4, uni_5, uni_7, uni_9)

R2_vals <- unlist(lapply(beta_res_list, function(x) x[,"R2"][3]))
p_vals <- unlist(lapply(beta_res_list, function(x) x[,"Pr(>F)"][3]))
beta_res_df <- as.data.frame(cbind(R2_vals, p_vals))
rownames(beta_res_df) <- as.character(lapply(beta_res_list, function(x) rownames(x)[3]))

beta_res_df <- beta_res_df[order(beta_res_df$R2_vals, decreasing = F),]


#varying color for p values
#n of group size
par(mar=c(4,14,1,1))
barplot(beta_res_df$R2_vals, horiz = T, las=1, col = beta_res_df$p_vals < 0.05, names.arg = rownames(beta_res_df))
mtext("R2 % variance explained", line=2.5, side=1)


#filename <- "../figures/barplot fecal mx weighted manhattan manuscript.png"
#png(filename, width = 8*330, height = 5*330, res = 330)
#par(mar=c(4,14,1,1))
#barplot(beta_res_df$R2_vals, horiz = T, las=1, col = beta_res_df$p_vals < 0.05, names.arg = rownames(beta_res_df))
#mtext("R2 % variance explained", line=2.5, side=1)
#dev.off()


#write.csv(beta_res_df, file = "../data/beta_res manhattan stool metabolome cor for Age and Bristol Score.csv")


```


Adjust for more variables, BMI

```{r, echo=FALSE, results='hide',message=FALSE}

#looks the same on all and subsetted by prevalence data
beta_div <- as.matrix(vegdist(t(mx_data), method = "manhattan"))


set.seed(1)
# Permanova - Distance based multivariate analysis of variance
uni_1a <- GUniFrac::adonis3(beta_div ~ Bristol_score_imputed + Age + IMDC, data = clin_meta_mx, permutations = 999)$aov.tab
uni_1b <- GUniFrac::adonis3(beta_div ~ Bristol_score_imputed + Age + hepatitis, data = clin_meta_mx, permutations = 999)$aov.tab
uni_1c <- GUniFrac::adonis3(beta_div ~ Bristol_score_imputed + Age + pneumonitis, data = clin_meta_mx, permutations = 999)$aov.tab
uni_1d <- GUniFrac::adonis3(beta_div ~ Bristol_score_imputed + Age + other_irAE, data = clin_meta_mx, permutations = 999)$aov.tab

beta_res_list <- list(uni_1a, uni_1b, uni_1c, uni_1d)

R2_vals <- unlist(lapply(beta_res_list, function(x) x[,"R2"][3]))
p_vals <- unlist(lapply(beta_res_list, function(x) x[,"Pr(>F)"][3]))
beta_res_df <- as.data.frame(cbind(R2_vals, p_vals))
rownames(beta_res_df) <- as.character(lapply(beta_res_list, function(x) rownames(x)[3]))

beta_res_df <- beta_res_df[order(beta_res_df$R2_vals, decreasing = F),]

#varying color for p values
#n of group size
par(mar=c(4,14,1,1))
barplot(beta_res_df$R2_vals, horiz = T, las=1, col = beta_res_df$p_vals < 0.05, names.arg = rownames(beta_res_df))
mtext("R2 % variance explained", line=2.5, side=1)

write.csv(beta_res_df, file = "../data/beta_res_df_adj manhattan stool metabolome cor Bristol_score_imputed Age.csv")



```


Look at correlation between water content and Bristol stool form scale (BSFS)

```{r, echo=FALSE, results='hide',message=FALSE}

water_content_vec_sub <- clin_meta_mx$water_content_perc
summary(water_content_vec_sub)
summary(as.numeric(clin_meta_mx$Bristol_score_curated))

cor(water_content_vec_sub, as.numeric(clin_meta_mx$Bristol_score_curated), use="na.or.complete", method = "spearman")
cor.test(water_content_vec_sub, as.numeric(clin_meta_mx$Bristol_score_curated), use="na.or.complete", method = "spearman")

#very significantly correlated as expected

#	Spearman's rank correlation rho
#data:  water_content_vec_sub and as.numeric(clin_meta_mx$Bristol_score_curated)
#S = 353462, p-value = 8.069e-09
#alternative hypothesis: true rho is not equal to 0
#sample estimates:
#      rho 
#0.4413515 


plot(water_content_vec_sub, as.numeric(clin_meta_mx$Bristol_score_curated), pch=16, las=1, 
     xlab="water content %", ylab="Bristol Stool Form Scale")


```


Function to replace outliers with median, on dataframe

```{r, echo=FALSE, results='hide',message=FALSE}

replace_outliers_with_median <- function(df, groups, n_sd=4, verbose=FALSE) {
  #metabolites assumed to be in rows, groups in columns
  
  #get the medians per group per metabolite
  median_values <- list()
  df_output <- df
  for (i in 1:nrow(df)) {
  x <- as.numeric(df[i,])
  median_values[[i]] <- aggregate(x[!is.na(x)], by=list(groups[!is.na(x)]), FUN=median)
  outlier_index <- which((x - mean(x)) / sd(x) > n_sd)
  if (length(outlier_index) > 0) {
    if  (verbose==FALSE) {
    cat("number of removed outliers out of:\n")
    print(length(outlier_index))
    print(length(x))
    }
    for (j in 1:length(outlier_index)) {
      na_group <- groups[outlier_index[j]]
      group_index <- which(groups == na_group)
      replace_val <- median_values[[i]]$x[median_values[[i]]$Group.1 == na_group]
      df_output[i,outlier_index[j]] <- replace_val
    }
  }
  }
  return(df_output)
}

```


Relevel 

```{r, echo=FALSE, results='hide',message=FALSE}

clin_meta_mx$IMDC <- relevel(factor(clin_meta_mx$IMDC), ref = "1")
clin_meta_mx$hepatitis <- relevel(factor(clin_meta_mx$hepatitis), ref = "1")
clin_meta_mx$pneumonitis <- relevel(factor(clin_meta_mx$pneumonitis), ref = "1")
clin_meta_mx$other_irAE <- relevel(factor(clin_meta_mx$other_irAE), ref = "1")

```


IMDC t.test and linear model corrected for Bristol_score_imputed and Age on normalized data

```{r, echo=FALSE, results='hide',message=FALSE}


mtbs_IMDC <- replace_outliers_with_median(mtbs, clin_meta_mx$IMDC, verbose = T)
mtbs_imp_IMDC <- replace_outliers_with_median(mtbs_imp, clin_meta_mx$IMDC, verbose = T)


p_list <- list()
lm_p_list <- list()
fc_list <- list()
fc_list2 <- list()
for (i in 1:nrow(mtbs)) {
  
  x <- mtbs_IMDC[i,]
  
  temp_test <- t.test(x ~ clin_meta_mx$IMDC)
  lm_temp_test <- lm(x ~ clin_meta_mx$IMDC + clin_meta_mx$Bristol_score_imputed + clin_meta_mx$Age)
  lm_temp_test_summary <- summary(lm_temp_test)
  
  lm_p_list[[i]] <- as.numeric(lm_temp_test_summary$coefficients[, "Pr(>|t|)"][2])
  p_list[[i]] <- as.numeric(temp_test$p.value)
  fc_list[[i]] <- log2(median(mtbs_imp_IMDC[i, clin_meta_mx$IMDC == "1"]) / median(mtbs_imp_IMDC[i, clin_meta_mx$IMDC == "0"]))
  fc_list2[[i]] <- as.numeric(temp_test$estimate[1] - temp_test$estimate[2])
  
}

res_df <- as.data.frame(cbind("t_p" = unlist(p_list), "p" = unlist(lm_p_list), "log2FC" = unlist(fc_list)))
rownames(res_df) <- rownames(mtbs)
res_df$q <- p.adjust(res_df$p, method = "fdr") 

res_df_IMDC <- res_df


plot(res_df$log2FC, -log10(res_df$t_p), las=2)
abline(h=c(-log10(0.05)))

plot(res_df$log2FC, -log10(res_df$p), las=2)
abline(h=c(-log10(0.05)))
abline(v=c(-1,1))

res_df[order(res_df$p, decreasing = F), ]

res_df_IMDC <- res_df


```


IMDC Pathway enrichment on p<0.05 

```{r, echo=FALSE, results='hide',message=FALSE}

matched_df <- as.data.frame(HMDB_pathways_matched_mtbs)

sig_df <- res_df_IMDC[which(res_df_IMDC$p < 0.05),]
names_sig <- rownames(sig_df)

HMDB_pathways_matched_mtbs[rownames(sig_df),]

#taxonomy_class very significant with Diazines 
enricher_output <- enricher_function(names_sig, level="taxonomy_class", matched_df=matched_df, pvalueCutoff=0.25, qvalueCutoff = 0.25)
#enricher_output

#subclass looks better 
enricher_output_IMDC <- enricher_function(names_sig, level="taxonomy_subclass", matched_df=matched_df, pvalueCutoff=0.25, qvalueCutoff = 0.25)
enricher_output_IMDC

enricher_output_IMDC$res_df
#q-value 0.004069915 for Pyrimidines and pyrimidine derivatives, p.adjust 0.005412987 as the only category <0.1
#Cytosine/Orotic.acid/Uracil with Orotic.acid having the largest effect size



```


plot boxplots for the significant metabolites IMDC

```{r, echo=FALSE, results='hide',message=FALSE}


res_df_IMDC_sig <- res_df_IMDC[res_df_IMDC$p < 0.05 & abs(res_df_IMDC$log2FC) > 1,]
sig_metabolites <- rownames(res_df_IMDC_sig)


clin_meta_mx$IMDC <- relevel(factor(clin_meta_mx$IMDC), ref = "0")
groups <- clin_meta_mx$IMDC
sel_colors <- adjustcolor(c("darkgrey", "#FEB24C"), alpha.f = 0.9)


#pdfs for all
pdf(paste("../figures/", "stool metabolites IMDC", ".pdf", sep=""), width=3, height=4)    

for (i in 1:length(sig_metabolites)) {
  
  temp_name <- sig_metabolites[i]
  x <- as.numeric(mx_data[temp_name,])

  par(mar=(c(2.5,4.5,2,0.2)))
  boxplot(x ~ groups, varwidth=F, las=1, xlab="", ylab="", ylim=range(x), outline=F, border=sel_colors, 
          horizontal=F, col="white", names=c("no IMDC", "IMDC"))
  parusr <- par('usr')
  scale_temp <- 1.17
  segments(1, parusr[4]/scale_temp, 2, parusr[4]/scale_temp) 
  scale_temp <- 1.12
  
  temp_p <- res_df_IMDC_sig[temp_name,"p"]
  
  text(1.5, parusr[4]/scale_temp, paste("p=", round(temp_p,4), sep=""), cex=0.8, font=1)

  mtext("uMole/gram stool", 2, line=3.5)
  mtext(temp_name, side=3, line=0, adj=0, cex.main=1, font.main=1)
  stripchart(x ~ groups, vertical = T, method = "jitter", jitter= 0.25, add = TRUE, pch=16, col=sel_colors, cex=0.8)
}
dev.off()


```


hepatitis t.test and linear model corrected for Bristol_score_imputed and Age on normalized data

```{r, echo=FALSE, results='hide',message=FALSE}


mtbs_hepatitis <- replace_outliers_with_median(mtbs, clin_meta_mx$hepatitis, verbose = T)
mtbs_imp_hepatitis <- replace_outliers_with_median(mtbs_imp, clin_meta_mx$hepatitis, verbose = T)


p_list <- list()
lm_p_list <- list()
fc_list <- list()
fc_list2 <- list()
for (i in 1:nrow(mtbs)) {
  
  x <- mtbs_hepatitis[i,]
  
  temp_test <- t.test(x ~ clin_meta_mx$hepatitis)
  lm_temp_test <- lm(x ~ clin_meta_mx$hepatitis + clin_meta_mx$Bristol_score_imputed + clin_meta_mx$Age)
  lm_temp_test_summary <- summary(lm_temp_test)
  
  lm_p_list[[i]] <- as.numeric(lm_temp_test_summary$coefficients[, "Pr(>|t|)"][2])
  p_list[[i]] <- as.numeric(temp_test$p.value)
  
  fc_list[[i]] <- log2(median(mtbs_imp_hepatitis[i, clin_meta_mx$hepatitis == "1"]) / median(mtbs_imp_hepatitis[i, clin_meta_mx$hepatitis == "0"]))
  fc_list2[[i]] <- as.numeric(temp_test$estimate[1] - temp_test$estimate[2])
  
}

res_df <- as.data.frame(cbind("t_p" = unlist(p_list), "p" = unlist(lm_p_list), "log2FC" = unlist(fc_list)))
rownames(res_df) <- rownames(mtbs)
res_df$q <- p.adjust(res_df$p, method = "fdr") 

res_df_hepatitis <- res_df

res_df_hepatitis[order(as.numeric(res_df_hepatitis$log2FC)),]
res_df_hepatitis[order(as.numeric(res_df_hepatitis$p), decreasing = F),]
res_df_hepatitis[as.numeric(res_df_hepatitis$p) < 0.05,]


plot(res_df$log2FC, -log10(res_df$p), las=2)
abline(h=c(-log10(0.05)))
abline(v=c(-1,1))


```

Hepatitis Pathway enrichment on p<0.05 

```{r, echo=FALSE, results='hide',message=FALSE}

sig_df <- res_df_hepatitis[which(res_df_hepatitis$p < 0.05),]
names_sig <- rownames(sig_df)

HMDB_pathways_matched_mtbs[rownames(sig_df),]

#taxonomy_class Indoles and derivatives p.adjust 0.0346 with 2 metabolites
enricher_output <- enricher_function(names_sig, level="taxonomy_class", matched_df=matched_df, pvalueCutoff=0.25, qvalueCutoff = 0.25)
#enricher_output

#subclass looks better 
enricher_output_Hepatitis <- enricher_function(names_sig, level="taxonomy_subclass", matched_df=matched_df, pvalueCutoff=0.25, qvalueCutoff = 0.25)
enricher_output_Hepatitis

enricher_output_Hepatitis$res_df
#no pathways (taxonomy_subclass) with more than 1 metabolite. enricher_output_Hepatitis$res_df 0.197


```


plot boxplots for the significant metabolites Hepatitis

```{r, echo=FALSE, results='hide',message=FALSE}


res_df_hepatitis_sig <- res_df_hepatitis[res_df_hepatitis$p < 0.05 & abs(res_df_hepatitis$log2FC) > 1,]
sig_metabolites <- rownames(res_df_hepatitis_sig)


clin_meta_mx$hepatitis <- relevel(factor(clin_meta_mx$hepatitis), ref = "0")
groups <- clin_meta_mx$hepatitis
sel_colors <- adjustcolor(c("darkgrey", "#FEB24C"), alpha.f = 0.9)


#pdfs for all
pdf(paste("../figures/", "stool metabolites hepatitis", ".pdf", sep=""), width=3, height=4)    

for (i in 1:length(sig_metabolites)) {
  
  temp_name <- sig_metabolites[i]
  x <- as.numeric(mx_data[temp_name,])

  par(mar=(c(2.5,4.5,2,0.2)))
  boxplot(x ~ groups, varwidth=F, las=1, xlab="", ylab="", ylim=range(x), outline=F, border=sel_colors, 
          horizontal=F, col="white", names=c("no hepatitis", "hepatitis"))
  parusr <- par('usr')
  scale_temp <- 1.17
  segments(1, parusr[4]/scale_temp, 2, parusr[4]/scale_temp) 
  scale_temp <- 1.12
  
  temp_p <- res_df_hepatitis_sig[temp_name,"p"]
  
  text(1.5, parusr[4]/scale_temp, paste("p=", round(temp_p,4), sep=""), cex=0.8, font=1)

  mtext("uMole/gram stool", 2, line=3.5)
  mtext(temp_name, side=3, line=0, adj=0, cex.main=1, font.main=1)
  stripchart(x ~ groups, vertical = T, method = "jitter", jitter= 0.25, add = TRUE, pch=16, col=sel_colors, cex=0.8)
}
dev.off()


```


pneumonitis t.test and linear model corrected for Bristol_score_imputed and Age on normalized data

```{r, echo=FALSE, results='hide',message=FALSE}

mtbs_pneu <- replace_outliers_with_median(mtbs, clin_meta_mx$pneumonitis, verbose = T)
mtbs_imp_pneu <- replace_outliers_with_median(mtbs_imp, clin_meta_mx$pneumonitis, verbose = T)


p_list <- list()
lm_p_list <- list()
fc_list <- list()
fc_list2 <- list()
for (i in 1:nrow(mtbs)) {
  x <- as.numeric(mtbs_pneu[i,])
  temp_test <- t.test(x ~ clin_meta_mx$pneumonitis)
  lm_temp_test <- lm(x ~ clin_meta_mx$pneumonitis + clin_meta_mx$Bristol_score_imputed + clin_meta_mx$Age)
  lm_temp_test_summary <- summary(lm_temp_test)
  
  lm_p_list[[i]] <- as.numeric(lm_temp_test_summary$coefficients[, "Pr(>|t|)"][2])
  p_list[[i]] <- as.numeric(temp_test$p.value)
  fc_list[[i]] <- log2(median(mtbs_imp_pneu[i, clin_meta_mx$pneumonitis == "1"]) / median(mtbs_imp_pneu[i, clin_meta_mx$pneumonitis == "0"]))
  fc_list2[[i]] <- as.numeric(temp_test$estimate[1] - temp_test$estimate[2])
  
}

res_df <- as.data.frame(cbind("t_p" = unlist(p_list), "p" = unlist(lm_p_list), "log2FC" = unlist(fc_list)))
rownames(res_df) <- rownames(mtbs)
res_df$q <- p.adjust(res_df$p, method = "fdr") 

res_df_pneumonitis <- res_df

res_df_pneumonitis[order(as.numeric(res_df_pneumonitis$log2FC)),]
res_df_pneumonitis[order(as.numeric(res_df_pneumonitis$p), decreasing = F),]
res_df_pneumonitis[as.numeric(res_df_pneumonitis$p) < 0.05,]


plot(res_df$log2FC, -log10(res_df$p), las=2)
abline(h=c(-log10(0.05)))
abline(v=c(-1,1))


```


Pneumonitis Pathway enrichment on p<0.05 

```{r, echo=FALSE, results='hide',message=FALSE}

sig_df <- res_df_pneumonitis[which(res_df_pneumonitis$p < 0.05),]
names_sig <- rownames(sig_df)

HMDB_pathways_matched_mtbs[rownames(sig_df),]

#taxonomy_class nothing
enricher_output <- enricher_function(names_sig, level="taxonomy_class", matched_df=matched_df, pvalueCutoff=0.25, qvalueCutoff = 0.25)
enricher_output

#subclass looks better overall
enricher_output_Pneumonitis <- enricher_function(names_sig, level="taxonomy_subclass", matched_df=matched_df, pvalueCutoff=0.25, qvalueCutoff = 0.25)
enricher_output_Pneumonitis


enricher_output_Pneumonitis$res_df
#Glycosphingolipids (taxonomy_subclass) with 5 metabolites p.adjust 0.13; so not significant either


```


plot boxplots for the significant metabolites Pneu

```{r, echo=FALSE, results='hide',message=FALSE}

res_df_pneumonitis_sig <- res_df_pneumonitis[res_df_pneumonitis$p < 0.05 & abs(res_df_pneumonitis$log2FC) > 1,]
sig_metabolites <- rownames(res_df_pneumonitis_sig)


clin_meta_mx$pneumonitis <- relevel(factor(clin_meta_mx$pneumonitis), ref = "0")
groups <- clin_meta_mx$pneumonitis
sel_colors <- adjustcolor(c("darkgrey", "#FEB24C"), alpha.f = 0.9)


#pdfs for all
pdf(paste("../figures/", "stool metabolites pneu", ".pdf", sep=""), width=3, height=4)    

for (i in 1:length(sig_metabolites)) {
  
  temp_name <- sig_metabolites[i]
  x <- as.numeric(mx_data[temp_name,])

  par(mar=(c(2.5,4.5,2,0.2)))
  boxplot(x ~ groups, varwidth=F, las=1, xlab="", ylab="", ylim=range(x), outline=F, border=sel_colors, 
          horizontal=F, col="white", names=c("no Pneu", "Pneu"))
  parusr <- par('usr')
  scale_temp <- 1.17
  segments(1, parusr[4]/scale_temp, 2, parusr[4]/scale_temp) 
  scale_temp <- 1.12
  
  temp_p <- res_df_pneumonitis_sig[temp_name,"p"]
  
  text(1.5, parusr[4]/scale_temp, paste("p=", round(temp_p,4), sep=""), cex=0.8, font=1)

  mtext("uMole/gram stool", 2, line=3.5)
  mtext(temp_name, side=3, line=0, adj=0, cex.main=1, font.main=1)
  stripchart(x ~ groups, vertical = T, method = "jitter", jitter= 0.25, add = TRUE, pch=16, col=sel_colors, cex=0.8)
}
dev.off()


```


other_irAE t.test and linear model corrected for Bristol_score_imputed and Age on normalized data

```{r, echo=FALSE, results='hide',message=FALSE}

mtbs_other <- replace_outliers_with_median(mtbs, clin_meta_mx$other_irAE, verbose = T)
mtbs_imp_other <- replace_outliers_with_median(mtbs_imp, clin_meta_mx$other_irAE, verbose = T)


p_list <- list()
lm_p_list <- list()
fc_list <- list()
fc_list2 <- list()

for (i in 1:nrow(mtbs)) {
  
  x <- mtbs_other[i,]
  
  temp_test <- t.test(x ~ clin_meta_mx$other_irAE)
  lm_temp_test <- lm(x ~ clin_meta_mx$other_irAE + clin_meta_mx$Bristol_score_imputed + clin_meta_mx$Age)
  lm_temp_test_summary <- summary(lm_temp_test)
  
  lm_p_list[[i]] <- as.numeric(lm_temp_test_summary$coefficients[, "Pr(>|t|)"][2])
  p_list[[i]] <- as.numeric(temp_test$p.value)
  fc_list[[i]] <- log2(median(mtbs_imp_other[i, clin_meta_mx$other_irAE == "1"]) / median(mtbs_imp_other[i, clin_meta_mx$other_irAE == "0"]))
  fc_list2[[i]] <- as.numeric(temp_test$estimate[1] - temp_test$estimate[2])
}

res_df <- as.data.frame(cbind("t_p" = unlist(p_list), "p" = unlist(lm_p_list), "log2FC" = unlist(fc_list)))
rownames(res_df) <- rownames(mtbs)
res_df$q <- p.adjust(res_df$p, method = "fdr") 

res_df_other_irAE <- res_df

res_df_other_irAE[order(as.numeric(res_df_other_irAE$log2FC)),]
res_df_other_irAE[order(as.numeric(res_df_other_irAE$p), decreasing = F),]
res_df_other_irAE[as.numeric(res_df_other_irAE$p) < 0.05,]


#more significant metabolites but nothing beyond 1 log2(FC)
plot(res_df$log2FC, -log10(res_df$p), las=2)
abline(h=c(-log10(0.05)))



```


other_irAE Pathway enrichment on p<0.05 

```{r, echo=FALSE, results='hide',message=FALSE}

sig_df <- res_df_other_irAE[which(res_df_other_irAE$p < 0.05),]
names_sig <- rownames(sig_df)

HMDB_pathways_matched_mtbs[rownames(sig_df),]

#taxonomy_class nothing
enricher_output <- enricher_function(names_sig, level="taxonomy_class", matched_df=matched_df, pvalueCutoff=0.25, qvalueCutoff = 0.25)
enricher_output

#subclass looks better overall
enricher_output_other_irAE <- enricher_function(names_sig, level="taxonomy_subclass", matched_df=matched_df, pvalueCutoff=0.25, qvalueCutoff = 0.25)
enricher_output_other_irAE


enricher_output_other_irAE$res_df
#no p.adjust <0.2


```


Combine the results

```{r, echo=FALSE, message=FALSE}

comb_res_list <- list(res_df_IMDC, res_df_hepatitis, res_df_pneumonitis, res_df_other_irAE)
names(comb_res_list) <- c("IMDC", "hepatitis", "pneumonitis", "other irAE")

df.p <- as.data.frame(do.call(cbind, lapply(comb_res_list, function(x) x$p)), row.names = rownames(res_df_IMDC))
df.fc <- as.data.frame(do.call(cbind, lapply(comb_res_list, function(x) x$log2FC)), row.names = rownames(res_df_IMDC))


#also combine the enricher_output_other_irAE
cutoff <- 0.05
enricher_output_IMDC$res_df[enricher_output_IMDC$res_df$p.adjust < cutoff,]
#enricher_output_Hepatitis$res_df[enricher_output_Hepatitis$res_df$p.adjust < cutoff,]
#enricher_output_Pneumonitis$res_df[enricher_output_Pneumonitis$res_df$p.adjust < cutoff,]
#enricher_output_other_irAE$res_df[enricher_output_other_irAE$res_df$p.adjust < cutoff,]

#                                             ID  Description GeneRatio BgRatio RichFactor
#Pyrimidines and pyrimidine derivatives      3/18   3/186          1
#FoldEnrichment   zScore    pvalue    p.adjust      qvalue         geneID                   Count
#      10.33333 5.320339 0.0007732838 0.005412987 0.004069915 Cytosine/Orotic.acid/Uracil     3


```

```{r, echo=FALSE, message=FALSE}

#assign some colors (consistent with the metagenomics code file)
col <- c(brewer.pal(12,'Set3'),brewer.pal(12,'Paired'),brewer.pal(9,'Set1'),brewer.pal(8,'Set2'))
#length(col) #41

###### only do one time here; no need to repeat for genus level below
# Generate 187 unique random colors
n <- 187
colors <- character(0)

while (length(colors) < n) {
  remaining <- n - length(colors)
  set.seed(8)
  new_colors <- sprintf("#%06x", sample(0:16777215, remaining, replace = TRUE))
  colors <- unique(c(colors, new_colors))
  colors <- colorspace::darken(colors, amount = 0.2)
}
colors <- colors[1:n]  # Ensure exact length

```


Heatmap of irAE filtered by significance and fold change

```{r, echo=FALSE, message=FALSE}


grp.names <- colnames(df.p)
grp.labels <- colnames(df.p)

#p and fc cutoff
cutoff <- 0.05; q.cut1 = 0.001; q.cut2=0.01;q.cut3=0.05
log2FC_cutoff <- 1

#P_sig is simply one where a row contains a number 1 below cutoff
P.sig <- df.p[rowSums(df.p <= cutoff) >= 1,]
FC.sig <- df.fc[rowSums(df.p <= cutoff) >= 1,]

#select FC cutoff as well
FC.sig <- FC.sig[rowSums(abs(FC.sig) >= log2FC_cutoff) >= 1,]
P.sig <- P.sig[rownames(FC.sig),]

lgd_sig_p = Legend(pch = c("***", "**","*"), type = "points", labels = c("<0.001", "<0.01", "<0.05"), background="white", title = "p")

FC.sig <- as.matrix(t(FC.sig))
P.sig <- as.matrix(t(P.sig))


#table(HMDB_pathways_matched_mtbs[colnames(FC.sig), "taxonomy_subclass"])
#               Amino acids, peptides, and analogues            Benzoic acids and derivatives Indolyl carboxylic acids and derivatives 
#                                       3                                        1                                        1 
#                                   Methoxyphenols           Purines and purine derivatives   Pyrimidines and pyrimidine derivatives 
#                                       1                                        1                                        2 
#             Tricarboxylic acids and derivatives              Tryptamines and derivatives 
#                                       1                                        1 

vec <- HMDB_pathways_matched_mtbs[colnames(FC.sig), "taxonomy_subclass"]
label <- vec
vec_test <- !vec %in% names(which(table(vec) > 1))
label[vec_test] <- "Other"

col <- c("lightgrey", "#9080BA", "#C275DA")
names(col) <- unique(label)


#names, make wider and add names
ha_combined <- HeatmapAnnotation(
  Pathway = label,
  simple_anno_size = unit(0.4, "inch"),
  annotation_height = unit(2, "inch"),
  border = F,
  col = list(Pathway = col),
  show_legend = T,
  annotation_legend_param = list(
    title_gp = gpar(fontsize = 10, fontface = "bold"),  # Legend title size
    labels_gp = gpar(fontsize = 12)                    # Legend label size
  )
)



test_df <- FC.sig
test_df["IMDC", label == "Pyrimidines and pyrimidine derivatives"] <- "draw"


#replace dot by space in rownames
colnames(FC.sig) <- gsub("\\.", " ", colnames(FC.sig))

#make the scale even bidirectional
scale_max <- max(c(abs(min(FC.sig)), max(FC.sig)))

heatmap_full <- Heatmap(FC.sig, name = "log2(fold-change)", 
        heatmap_legend_param = list(direction = "vertical"),
        col = colorRamp2(c(-scale_max, 0, scale_max), 
                           c("#2171B5",'white', "#FEB24C")),
        column_gap = unit(1, "mm"), 
        border = TRUE,
        column_names_rot = 40,
        
        top_annotation = ha_combined,
        
        cell_fun = function(j, i, x, y, w, h, fill) {
          if(P.sig[i,j] <= q.cut1) {
            grid.text("***", x = x, y = y, hjust = 0.5, vjust = 0.7, gp = gpar(fontsize=12))
          }else if (P.sig[i,j] <= q.cut2) {
            grid.text("**", x = x, y = y, hjust = 0.5, vjust = 0.7, gp = gpar(fontsize=12))
          }else if (P.sig[i,j] <= q.cut3) {
            grid.text("*", x = x, y = y, hjust = 0.5, vjust = 0.7, gp = gpar(fontsize=12))
          }
          # Condition to draw an extra box around specific cells, pyrimidine pathway
          if (test_df[i, j] == "draw") {
            grid.rect(x = x, y = y, width = w * 0.95, height = h * 0.95,
              gp = gpar(col = "#C275DA", fill = NA, lwd = 3))  
          }
        },         
        rect_gp = gpar(col= "white"),
        show_column_names = T, show_row_names = T,
        show_column_dend = F,
        show_row_dend = F,
        cluster_rows = T,
        cluster_columns = T,
        column_names_gp = gpar(fontsize = 14),
        row_names_max_width = max_text_width(rownames(FC.sig), gp = gpar(fontsize = 20))
) 
heatmap_full

#packed_legends <- packLegend(lgd_sig_p, lgd_sig_fdr, direction = "vertical")
packed_legends <- packLegend(lgd_sig_p, direction = "vertical")

heatmap_plot2 <- draw(heatmap_full, padding = unit(c(0.1, 4, 0.1, 0.1), "cm"), annotation_legend_list = list(packed_legends), merge_legend=T) 
heatmap_plot2


filename <- "../figures/mx ICI irAE heatmap log2_above1_w_pathway.png"
png(filename, width = 10*330, height = 3.5*330, res = 330)
heatmap_plot2
dev.off()



```


Plot individual metabolites IMDC

```{r, echo=FALSE, message=FALSE}

res_df_IMDC_sig <- res_df_IMDC[res_df_IMDC$p < 0.05 & abs(res_df_IMDC$log2FC) > 1,]
sig_metabolites <- rownames(res_df_IMDC_sig)

#"4-Hydroxybenzoic.acid" "Alpha-aminobutyric.acid" "Cytosine" "Indolelactic.acid" "Orotic.acid" "Xanthine"  

clin_meta_mx$IMDC <- relevel(factor(clin_meta_mx$IMDC), ref = "0")
groups <- clin_meta_mx$IMDC
sel_colors <- adjustcolor(c("darkgrey", "#FEB24C"), alpha.f = 0.9)




#"4-Hydroxybenzoic.acid"
x <- as.numeric(mx_data["4-Hydroxybenzoic.acid",])
range(x)

plotname <- "Hydroxybenzoic.acid manuscript"
png(paste("../figures/", plotname, ".png", sep=""), width=3*250, height=4*250, res=300)    
par(mar=(c(2.5,4.5,2,0.2)))
#manual ylim, all shown
boxplot(x ~ groups, varwidth=F, las=1, xlab="", ylab="", outline=F, ylim=c(0, 0.1), border=sel_colors, 
          horizontal=F, col="white", names=c("no IMDC", "IMDC"))
parusr <- par('usr')
scale_temp <- 1.13
segments(1, parusr[4]/scale_temp, 2, parusr[4]/scale_temp) 
scale_temp <- 1.08
text(1.5, parusr[4]/scale_temp, paste("p=", round(res_df_IMDC_sig["4-Hydroxybenzoic.acid", "p"],4), sep=""), cex=0.8, font=1)

mtext("uMole/gram stool", 2, line=3.5)
mtext("4-Hydroxybenzoic acid", side=3, line=0, at=0.25, adj=0, cex.main=1, font.main=1)
stripchart(x ~ groups, vertical = T, method = "jitter", jitter= 0.25, add = TRUE, pch=16, col=sel_colors, cex=0.8)
dev.off()




#"Alpha-aminobutyric.acid"
x <- as.numeric(mx_data["Alpha-aminobutyric.acid",])
range(x) #0.00338432 23.00000000

plotname <- "Alpha-aminobutyric.acid manuscript"
png(paste("../figures/", plotname, ".png", sep=""), width=3*250, height=4*250, res=300)    
par(mar=(c(2.5,4.5,2,0.2)))
#manual ylim, 1 point in non-IMDC not shown
boxplot(x ~ groups, varwidth=F, las=1, xlab="", ylab="", outline=F, ylim=c(0,4.75), border=sel_colors, 
          horizontal=F, col="white", names=c("no IMDC", "IMDC"))
parusr <- par('usr')
scale_temp <- 1.13
segments(1, parusr[4]/scale_temp, 2, parusr[4]/scale_temp) 
scale_temp <- 1.08
text(1.5, parusr[4]/scale_temp, paste("p=", round(res_df_IMDC_sig["Alpha-aminobutyric.acid", "p"],4), sep=""), cex=0.8, font=1)

mtext("uMole/gram stool", 2, line=3.5)
mtext("Alpha-aminobutyric acid", side=3, line=0, at=0.15, adj=0, cex.main=1, font.main=1)
stripchart(x ~ groups, vertical = T, method = "jitter", jitter= 0.25, add = TRUE, pch=16, col=sel_colors, cex=0.8)
dev.off()




#Cytosine
x <- as.numeric(mx_data["Cytosine",])
range(x) #0.0000000 0.1465088

plotname <- "Cytosine manuscript"
png(paste("../figures/", plotname, ".png", sep=""), width=3*250, height=4*250, res=300)    
par(mar=(c(2.5,4.5,2,0.2)))
boxplot(x ~ groups, varwidth=F, las=1, xlab="", ylab="", ylim=c(0,0.17), outline=F, border=sel_colors, 
          horizontal=F, col="white", names=c("no IMDC", "IMDC"))
parusr <- par('usr')
scale_temp <- 1.17
segments(1, parusr[4]/scale_temp, 2, parusr[4]/scale_temp) 
scale_temp <- 1.12
text(1.5, parusr[4]/scale_temp, paste("p=", round(res_df_IMDC_sig["Cytosine", "p"],4), sep=""), cex=0.8, font=1)

mtext("uMole/gram stool", 2, line=3.5)
mtext("Cytosine", side=3, line=0, adj=0, cex.main=1, font.main=1)
stripchart(x ~ groups, vertical = T, method = "jitter", jitter= 0.25, add = TRUE, pch=16, col=sel_colors, cex=0.8)
dev.off()




#Indolelactic.acid
x <- as.numeric(mx_data["Indolelactic.acid",])
range(x) #0.00000000 0.08167742

plotname <- "Indolelactic acid manuscript"
png(paste("../figures/", plotname, ".png", sep=""), width=3*250, height=4*250, res=300)    
par(mar=(c(2.5,4.5,2,0.2)))
#manual ylim, 1 point in non-IMDC not shown
boxplot(x ~ groups, varwidth=F, las=1, xlab="", ylab="", outline=F, ylim=c(0, 0.06), border=sel_colors, 
          horizontal=F, col="white", names=c("no IMDC", "IMDC"))
parusr <- par('usr')
scale_temp <- 1.13
segments(1, parusr[4]/scale_temp, 2, parusr[4]/scale_temp) 
scale_temp <- 1.08
text(1.5, parusr[4]/scale_temp, paste("p=", round(res_df_IMDC_sig["Indolelactic.acid", "p"],4), sep=""), cex=0.8, font=1)

mtext("uMole/gram stool", 2, line=3.5)
mtext("Indolelactic acid", side=3, line=0, adj=0, cex.main=1, font.main=1)
stripchart(x ~ groups, vertical = T, method = "jitter", jitter= 0.25, add = TRUE, pch=16, col=sel_colors, cex=0.8)
dev.off()





#Orotic acid
x <- as.numeric(mx_data["Orotic.acid",])

#note 1 sample is outside the plot area
plotname <- "Orotic acid manuscript"
png(paste("../figures/", plotname, ".png", sep=""), width=3*250, height=4*250, res=300)    
par(mar=(c(2.5,4.5,2,0.2)))
#manual ylim, 1 points in non-IMDC not shown
boxplot(x ~ groups, varwidth=F, las=1, xlab="", ylab="", ylim=c(0,2), outline=F, border=sel_colors, 
          horizontal=F, col="white", names=c("no IMDC", "IMDC"))
parusr <- par('usr')
scale_temp <- 1.17
segments(1, parusr[4]/scale_temp, 2, parusr[4]/scale_temp) 
scale_temp <- 1.12
text(1.5, parusr[4]/scale_temp, paste("p=", round(res_df_IMDC_sig["Orotic.acid", "p"],4), sep=""), cex=0.8, font=1)

mtext("uMole/gram stool", 2, line=3.5)
mtext("Orotic acid", side=3, line=0, adj=0, cex.main=1, font.main=1)
stripchart(x ~ groups, vertical = T, method = "jitter", jitter= 0.25, add = TRUE, pch=16, col=sel_colors, cex=0.8)
dev.off()




#Xanthine
x <- as.numeric(mx_data["Xanthine",])
range(x) #0.07064993 10.83350101

plotname <- "Xanthine manuscript"
png(paste("../figures/", plotname, ".png", sep=""), width=3*250, height=4*250, res=300)    
par(mar=(c(2.5,4.5,2,0.2)))
boxplot(x ~ groups, varwidth=F, las=1, xlab="", ylab="", ylim=c(0,11), outline=F, border=sel_colors, 
          horizontal=F, col="white", names=c("no IMDC", "IMDC"))
parusr <- par('usr')
scale_temp <- 1.17
segments(1, parusr[4]/scale_temp, 2, parusr[4]/scale_temp) 
scale_temp <- 1.12
text(1.5, parusr[4]/scale_temp, paste("p=", round(res_df_IMDC_sig["Xanthine", "p"],4), sep=""), cex=0.8, font=1)

mtext("uMole/gram stool", 2, line=3.5)
mtext("Xanthine", side=3, line=0, adj=0, cex.main=1, font.main=1)
stripchart(x ~ groups, vertical = T, method = "jitter", jitter= 0.25, add = TRUE, pch=16, col=sel_colors, cex=0.8)
dev.off()





```


correlate metabolites with each other

```{r, echo=FALSE, message=FALSE}

library(Hmisc)

res <- rcorr(as.matrix(t(mtbs)))  # mtbs: your metabolites dataframe
cor_matrix <- res$r
p_matrix <- res$P

flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    var1 = rownames(cormat)[row(cormat)[ut]],
    var2 = colnames(cormat)[col(cormat)[ut]],
    cor  = cormat[ut],
    p    = pmat[ut]
  )
}

cor_long <- flattenCorrMatrix(cor_matrix, p_matrix)
cor_long <- cor_long[order(abs(cor_long$cor), decreasing = T),]

sig_cor <- subset(cor_long, p < 0.05 & abs(cor) > 0.25)
sig_cor

sig_cor[sig_cor$var1 == "Orotic.acid" | sig_cor$var2 == "Orotic.acid",]


#malic acid’s role is to provide the carbon skeleton (via aspartate) necessary for the de novo synthesis of the pyrimidine ring, which is assembled into orotic acid before being converted into nucleotides

#would increasing Malate production capacity increase orotic acid?
plot(mtbs["Orotic.acid",], mtbs["Malic.acid",]) #interesting but not amazing
plot(mtbs["Orotic.acid",], mtbs["Xanthine",])
plot(mtbs["Orotic.acid",], mtbs["Dimethylamine",])


sig_cor[sig_cor$var1 == "Butyric.acid" | sig_cor$var2 == "Butyric.acid",]


```


Correlate LBP with metabolome results

```{r, echo=FALSE, message=FALSE}


comb_list <- list(mtbs)
lapply(comb_list, dim)

x <- as.numeric(clin_meta_mx$LBP_conc_ng_ml)
cor_res_list <- vector(mode = "list", length = length(comb_list))
for (k in 1:length(comb_list)) {
  y_data_sub <- comb_list[[k]]
  cor_list <- vector(mode = "list", length = nrow(y_data_sub))
  for (i in 1:nrow(y_data_sub)) {
    y <- as.numeric(y_data_sub[i,])
    cor_list[[i]][1] <- cor(x, y, use="na.or.complete", method="spearman") #there might be NAs in the data
    cor_list[[i]][2] <- suppressWarnings(cor.test(x, y, use="na.or.complete", method="spearman")$p.value)
  }
  cor_df <- as.data.frame(do.call(rbind, cor_list), stringsAsFactors=F)
  names(cor_df) <- c("cor", "p.val")
  rownames(cor_df) <- rownames(y_data_sub)
  #omit NAs
  cor_df <- cor_df[!is.na(cor_df$cor),]
  cor_df$fdr <- p.adjust(cor_df$p.val, "fdr")
  cor_df <- cor_df[order(abs(cor_df$cor), decreasing = T),]
  cor_res_list[[k]] <- cor_df
}

lapply(cor_res_list, dim)

head(cor_res_list[[1]], 20)


```
